\textit{[něco jako?]} Generating lambda term individuals using forgetful A*

### 2013-10-23 tex/evogp.tex
evogp tex

\newenvironment{todo}
{ ~\\[0.5em]
  {\color{red}\textbf{TODO}}
  \begin{easylist}[itemize]}
{ \end{easylist}
  ~}

@@

\subsection{Ramped half-and-half}

\subsection{Term}

Here we  %formally 
describe programming language, 
in which we generate individual programs --- so called \lterms.

@@

\textit{Function application} and 
\textit{$\lambda$-abstraction} are concepts
well known from common programming languages. 
For example in JavaScript 
$(M~N)$ translates to expression \texttt{$M$($N$)} and
$\lamb{x}{M}$ translates to expression \texttt{function($x$)\{return $M$;\}}.
In other words, the function application 
corresponds to the act of supplying a function 
with an argument and
the $\lambda$-abstraction is equivalent to 
\textit{anonymous function}.

\begin{todo}
 & asi říct o zkratkách notací
\end{todo}

\subsubsection{Reductions}
(... zmiňovat ? a pokud ano tak tady nebo radši v sekci která diskutuje 
normálnost vygenerovanejch termů a jejich zkrácení pomocí $eta$-redukce)
\subsubsection{Tree representations}
(... zmiňovat ?)

\subsection{Type}

A \lterm as described above
corresponds to a program expression with no type information
included. Now we will describe \textit{types} (or \textit{type terms}).
After putting those two pieces 
(\textit{\lterms} and \textit{types}) together 
we will get system called \textit{simply typed $\lambda$-calculus}.

\begin{definition}
Let $A$ be set of {\it atomic type names}. 
Then $\mathbb{T}$ is set of {\it types} inductively defined as follows.
\begin{align*}
\alpha      \in A  &\then   \alpha \in \T \\
\sigma,\tau \in \T &\then ( \sigma \ar  \tau ) \in \T 
\end{align*}~

\end{definition}

@@

\begin{todo}
 & asi říct o zkratce notace
\end{todo}

\subsection{Context}

\begin{definition} 
~
\begin{enumerate}
 \item 	Let $\Lambda$ be set of {\it \lterms}. 
	Let $\mathbb{T}$ be set of {\it types}.       
	A {\it statement} $M : \sigma$ is a pair 
	$(M,\sigma) \in \Lambda \times \mathbb{T}$.
	Statement $M : \sigma$ is vocalized as 
	{\it "$M$ has type $\sigma$"}.
	The term $M$ is called the {\it subject} of the 
	statement $M : \sigma$.
 \item A \textit{declaration} is a statement 
 $x : \sigma$ where $x \in V \cup C$.

 \item A \textit{context} 
 %(or \textit{basis}) 
 is set of declarations with distinct variables as subjects.
\end{enumerate}~
\end{definition}

\begin{todo}
   & říct že $T \cup F$ v Std GP je u nás $\Gamma$
   & popsat $\Gamma,x:\tau$ notaci
 a to že to automaticky implikuje že x neni obsažena v $\Gamma$,
 tzn je nová...
  & že píšem $x:\sigma \in \Gamma$ místo $(x,\sigma) \in \Gamma$
\end{todo}

\subsection{Simply typed $\lambda$-calculus}

\begin{definition}
A statement $M\colon\sigma$ is \textit{derivable from}
a context $\Gamma$ (notation 
\mbox{$\Gamma\vdash{}M\colon\sigma$}) 
if it can be produced by the following rules.
\begin{align*}
x : \sigma \in \Gamma &~\then~ \tur{\Gamma}{x}{\sigma}\\
\tur{\Gamma}{M}{\sigma \ar \tau}~,~\tur{\Gamma}{N}{\sigma} 
&~\then~ \tur{\Gamma}{(M~N)}{\tau}\\  
\tur{\Gamma,x:\sigma}{M}{\tau}
&~\then~ \tur{\Gamma}{\lamb{x}{M}}{\sigma \ar \tau} 
\end{align*}~
\end{definition}

@@

\begin{todo}
    & From this definition we can derive following rules...

	& citovat něco, asi barendrechta 
	(on to tam ale nějak moc nedokazuje) 
	případně mojí diplomku (tam to dokazuju)
\end{todo}

\subsection{Unfinished term}

@@

\begin{todo}
   & Vysvětlit k čemu je to dobrý.
\end{todo}

@@

Essential data structure of our term generating algorithm 
is priority queue of unfinished terms. 
Priority of an unfinished term is given by its size.
Initially, the queue contains only one unfinished term; 
$\ul{\tau}{\Gamma}$.\\

@@

Expansion of the selected unfinished leaf $L$ leads to creation of 
one or many (possibly zero) successors.
In this process, $L$ is replaced
by a new subterm defined by the following rules.
\begin{enumerate}
 \item If $L = \ul{\rho \ar \sigma}{\Gamma}$, 
       then $L$ is replaced by $\lamb{x}{\ul{\sigma}{\Gamma,x \colon \rho}}$.
       Thus this expansion results in exactly one successor.  
 \item If $L = \ul{\alpha}{\Gamma}$ where $\alpha$ is \textit{atomic} type,
       then for each 
       \mbox{$f : (\tau_1 \ar \dots \ar \tau_n \ar \alpha) \in \Gamma$}
       the unfinished leaf $L$ is replaced by 
       $(~f~(\tau_1,\Gamma)~\dots~(\tau_n,\Gamma)~)$.
       Thus this expansion results in many (possibly zero or one) successors.
\end{enumerate}

@@

If the queue becomes empty before the desired number $n$ of terms
is generated, then the initial unfinished term $\ul{\tau}{\Gamma}$ 
is inserted to the queue and the process continues.

@@

\subsubsection{Systematic strategy}
\subsubsection{Ramped half-and-half strategy}
\subsubsection{Geometric strategy}

~\\\\
\textit{
- Systematická strategie nezahodí nic (Pro ní generování 
  kolapsuje do A* algoritmu).\\
- Ramped-half-and-half zahodí vždy vše krom jediného.\\
- Geometrická strategie nezahodí s pravděpodobností q na hloubka UL \\
}

@@

Desired type of generated programs $\sigma$ and 
building blocks context $\Gamma$ are following.
\begin{align*}
\sigma = \Real \ar &\Real\\
\Gamma = \{
  (+)  &: \Real \ar \Real \ar \Real    ,
  (-)   : \Real \ar \Real \ar \Real    ,
  (*)   : \Real \ar \Real \ar \Real    ,
  rdiv  : \Real \ar \Real \ar \Real    ,\\
  sin  &: \Real \ar \Real              ,
  cos   : \Real \ar \Real              ,
  exp   : \Real \ar \Real              , 
  rlog  : \Real \ar \Real              \}
\end{align*}
where

\noindent\begin{minipage}{.5\linewidth}
\begin{align*}
rdiv(p,q) &= \begin{cases} 1 &\mbox{if } q = 0 \\
p/q & \mbox{otherwise } \end{cases}  
\end{align*}
\end{minipage}%
\begin{minipage}{.5\linewidth}
\begin{align*}
rlog(x) &= \begin{cases} 0 &\mbox{if } x = 0 \\
log(\vert x\vert) & \mbox{otherwise}. \end{cases}
\end{align*}
\end{minipage}

~\\\\
Fitness function is computed as follows

$$ fitness(f) =  \sum\limits_{i=1}^{20}{ \vert f(x_i)-y_i }\vert   $$

where $(x_i,y_i)$ are 20 data samples from $[-1,1]$, such that $y_i = f_t(x_i)$.\\

An individual $f$ such that $\vert f(x_i)-y_i \vert < 0.01 $ for all data samples is 
considered as a correct individual.

\subsection{Artificial ant}
...
\subsection{Even parity problem}
...
\section{Conclusions}
...

>>>>

\newenvironment{todo}
{~\\ {\color{red}\textbf{TODO}}
  \begin{easylist}[itemize]}
{ \end{easylist}}

@@

%\subsection{Term}

\Lets 
%Here we  %formally 
describe programming language, 
in which we generate individual programs --- so called \lterms.

@@

\textit{Function application} and 
\textit{$\lambda$-abstraction} are concepts
well known from common programming languages. 
For example in JavaScript 
$(M~N)$ translates to expression \texttt{$M$($N$)} and
$\lamb{x}{M}$ translates to expression \texttt{function($x$)\{return $M$;\}}.
In other words, the function application 
corresponds to the act of supplying a function 
with an argument and
the $\lambda$-abstraction is equivalent to 
\textit{anonymous function}.
%\begin{todo}
% & asi říct o zkratkách notací
%\end{todo}
%
%
%\subsubsection{Reductions}
%(... zmiňovat ? a pokud ano tak tady nebo radši v sekci která diskutuje 
%normálnost vygenerovanejch termů a jejich zkrácení pomocí $eta$-redukce)
%\subsubsection{Tree representations}
%(... zmiňovat ?)
%\subsection{Type}
A \lterm as described above
corresponds to a program expression with no type information
included. Now we will describe \textit{types} (or \textit{type terms}).
After putting those two pieces 
(\textit{\lterms} and \textit{types}) together 
we will get system called \textit{simply typed $\lambda$-calculus}.

\begin{definition}
Let $A$ be set of {\it atomic type names}. 
Then $\mathbb{T}$ is set of {\it types} inductively defined as follows.
\begin{align*}
\alpha      \in A  &\then   \alpha \in \T \\
\sigma,\tau \in \T &\then ( \sigma \ar  \tau ) \in \T 
\end{align*}
\end{definition}

@@

%\begin{todo}
% & asi říct o zkratce notace
%\end{todo}

%\subsection{Context}

\begin{definition}\begin{enumerate}
 \item 	Let $\Lambda$ be set of {\it \lterms}. 
	Let $\mathbb{T}$ be set of {\it types}.       
	A {\it statement} $M : \sigma$ is a pair 
	$(M,\sigma) \in \Lambda \times \mathbb{T}$.
	Statement $M : \sigma$ is vocalized as 
	{\it "$M$ has type $\sigma$"}.
	The term $M$ is called the {\it subject} of the 
	statement $M : \sigma$.
 \item A \textit{declaration} is a statement 
 $x : \sigma$ where $x \in V \cup C$.

 \item A \textit{context} 
 %(or \textit{basis}) 
 is set of declarations with distinct variables as subjects.
\end{enumerate}
\end{definition}

%\begin{todo}
%   & říct že $T \cup F$ v Std GP je u nás $\Gamma$
%   & popsat $\Gamma,x:\tau$ notaci
% a to že to automaticky implikuje že x neni obsažena v $\Gamma$,
% tzn je nová...
%  & že píšem $x:\sigma \in \Gamma$ místo $(x,\sigma) \in \Gamma$
%\end{todo}

%\subsection{Simply typed $\lambda$-calculus}

\begin{definition}
A statement $M\colon\sigma$ is \textit{derivable from}
a context $\Gamma$ (notation 
\mbox{$\Gamma\vdash{}M\colon\sigma$}) 
if it can be produced by the following rules.
\begin{align*}
x : \sigma \in \Gamma &~\then~ \tur{\Gamma}{x}{\sigma}\\
\tur{\Gamma}{M}{\sigma \ar \tau}~,~\tur{\Gamma}{N}{\sigma} 
&~\then~ \tur{\Gamma}{(M~N)}{\tau}\\  
\tur{\Gamma,x:\sigma}{M}{\tau}
&~\then~ \tur{\Gamma}{\lamb{x}{M}}{\sigma \ar \tau} 
\end{align*}
\end{definition}

@@

%\begin{todo}
%    & From this definition we can derive following rules...
%
%	& citovat něco, asi barendrechta 
%	(on to tam ale nějak moc nedokazuje) 
%	případně mojí diplomku (tam to dokazuju)
%\end{todo}

%\subsection{Unfinished term}

@@

%\begin{todo}
%   & Vysvětlit k čemu je to dobrý.
%\end{todo}

@@

Essential data structure of our algorithm 
is priority queue of unfinished terms. 
Priority of an unfinished term is given by its size.
At the beginning, the queue contains only one unfinished term; 
$\ul{\tau}{\Gamma}$. The search strategy $S$ also 
initializes its internal state (if it has one).\\

@@

Expansion of the selected unfinished leaf $L$ leads to creation of 
one or many (possibly zero) successors.
In this process, $L$ is replaced
by a new subterm defined by the following rules.
\begin{enumerate}
 \item If $L = \ul{\rho_1 \ar \dots \ar \rho_n \ar \alpha}{\Gamma}$,
 	   where $\alpha$ is atomic type and $n \geq 1$, 
       then $L$ is replaced by 
       $\lamb{x_1 \dots x_n}{\ul{\sigma}
       {\Gamma,x_1 \colon \rho_1,\dots,x_n \colon \rho_n}}$.
       Thus this expansion results in exactly one successor.  
 \item If $L = \ul{\alpha}{\Gamma}$ where $\alpha$ is \textit{atomic} type,
       then for each 
       \mbox{$f : (\tau_1 \ar \dots \ar \tau_m \ar \alpha) \in \Gamma$}
       the unfinished leaf $L$ is replaced by 
       $(~f~(\tau_1,\Gamma)~\dots~(\tau_m,\Gamma)~)$.
       Thus this expansion results in many (possibly zero or one) successors.
\end{enumerate}

@@

If the queue becomes empty before the desired number $n$ of terms
is generated, then the initial unfinished term $\ul{\tau}{\Gamma}$ 
is inserted to the queue, search strategy $S$
again initializes its internal state and the process continues.

@@

\begin{todo}
 & Okomentovat jak se počítá velikost a při tý příležitosti říct
   že A* heuristika se skrejvá v tom jak velký určíme unfinished leafs,
   že my používáme 1, ale že si de představit o dost sofistikovanější 
   verze.   
\end{todo}

\subsubsection{Systematic strategy}

If we use trivial strategy that returns all the inputs, 
then the algorithm systematically generates 
first $n$ smallest lambda terms in their
\textit{long normal form}.

\subsubsection{Ramped half-and-half strategy}

The internal state of this strategy consists of two variables.
It is the only one strategy described here that uses an internal state.

\begin{enumerate}
 \item \textit{isFull} - A boolean value, determining whether \textit{full}
                     or \textit{grow} method will be performed.
 \item \textit{d} - A integer value from $\setDots{2}{D_{init}}$, where 
                $D_{init}$ is predefined maximal depth (e.g. 6).                    
\end{enumerate}

This strategy returns precisely one randomly
(uniformly) selected  element from 
the \textit{selection subset} of input set
(or zero elements if the input set is empty). 
The \textit{selection subset} 
to select from is determined by $depth, d$ and $isFull$.
The $depth$ parameter is the depth (in the term tree) 
of the unfinished leaf that was expanded.
Those elements of input set whose newly added subtree contains one ore more 
unfinished leafs are regarded as \textit{non-terminals}, whereas 
those whose newly added subtree contains no unfinished leaf are regarded as 
\textit{terminals}.
If $depth = 0$, then the subset to select from is  
set of all \textit{non-terminals} of the input set.
If $depth = d$, then the subset to select from is
set of all \textit{terminals} of the input set.
In other cases of $depth$ it depends on value of $isFull$.
If $isFull = true$, then the subset to select from is 
set of all \textit{non-terminals} of the input set.
If $isFull = false$, then the subset to select from is 
the whole input set.

\begin{todo}
 & napsat o tom že ji používa koza [citace] a že je to standard,
 pokud je použita na gamu co splnuje closure podmínku, tak 
 generuje přesně stejně jako
 & ...This means that the queue always contains only one (or zero) state.
\end{todo}

\subsubsection{Geometric strategy}

We can see those two previous strategies as two extremes on the spectrum of 
possible strategies. 
\textit{Systematic strategy} filters no successor state thus performing
exhaustive search resulting in discovery of $n$ smallest terms in one run.
On the other hand, \textit{ramped half-and-half strategy} filters 
all but one successor states resulting in degradation of 
the priority queue into "fancy variable".
\textit{Geometric strategy} is simple yet fairly effective term generating 
strategy somewhere in the middle of this spectrum.
It is parameterized by parameter $q \in (0,1)$, its default well-performing 
value is $q = 0.75$.
For each element of the input set 
it is probabilistically decided whether
it will be returned or omitted. A probability $p$ of returning is
same for all elements, but depends on the $depth$, 
which is defined in the same way as in previous strategy. 
It is computed as follows.
$$ p = q^{depth} $$
This formula is motivated by idea that it is important to
explore all possible root symbols, but as the $depth$ 
increases it becomes less "dangerous" to omit 
an exploration branch. 
We can see this by considering that this strategy results in
somehow forgetful A* search.
With each omission we make the search space smaller. But with
increasing depth these omissions have smaller impact on the search space,
i.e., they cut out lesser portion of the search space.
Another slightly esoteric argument supporting this formula is that "root 
parts" of a program usually stand for more crucial parts
with radical impact on global behavior of a program, 
whereas "leaf parts" of a program usually
stand for less important local parts (e.g. constants).  
This strategy also plays nicely with the idea that 
"too big trees should be killed".

@@

Desired type of generated programs $\sigma$ and 
building blocks context $\Gamma$ are following.
\begin{align*}
\tau = \Real \ar &\Real\\
\Gamma = \{
  (+)  &: \Real \ar \Real \ar \Real    ,
  (-)   : \Real \ar \Real \ar \Real    ,
  (*)   : \Real \ar \Real \ar \Real    ,
  rdiv  : \Real \ar \Real \ar \Real    ,\\
  sin  &: \Real \ar \Real              ,
  cos   : \Real \ar \Real              ,
  exp   : \Real \ar \Real              , 
  rlog  : \Real \ar \Real              \}
\end{align*}
where
\noindent
\begin{minipage}{.5\linewidth}
\begin{align*}
rdiv(p,q) &= \begin{cases} 1 &\mbox{if } q = 0 \\
p/q & \mbox{otherwise } \end{cases}  
\end{align*}
\end{minipage}%
\begin{minipage}{.5\linewidth}
\begin{align*}
rlog(x) &= \begin{cases} 0 &\mbox{if } x = 0 \\
log(\vert x\vert) & \mbox{otherwise}. \end{cases}
\end{align*}
\end{minipage}
~\\
Fitness function is computed as follows.
$$ fitness(f) =  \sum\limits_{i=1}^{20}{ \vert f(x_i)-y_i }\vert   $$
where $(x_i,y_i)$ are 20 data samples from $[-1,1]$, such that $y_i = f_t(x_i)$.
An individual $f$ such that $\vert f(x_i)-y_i \vert < 0.01 $ for all data samples is 
considered as a correct individual.

\subsection{Artificial ant}

\textit{Artificial Ant} is another problem described
in \cite{koza92}. Objective of this problem is to 
find a control program for an artificial ant so
that it can find all food located on "Santa Fe" trail.
The Santa Fe trail lies on toroidal square grid.
The ant is in the upper left corner, facing right.
The ant is able to move forward, turn left, and sense if a food 
piece is ahead of him.

\begin{align*}
\tau = An&tAction\\
\Gamma = \{~~
  l    &: AntAction                              ,
  r    : AntAction                               ,
  m    : AntAction                               ,\\
  ifa  &: AntAction \ar AntAction \ar AntAction  ,\\
  p2   &: AntAction \ar AntAction \ar AntAction  ,\\
  p3   &: AntAction \ar AntAction \ar AntAction \ar AntAction  \}
\end{align*}

Action $l$ turns the ant left. 
Action $r$ turns the ant right.
Action $m$ moves the ant forward.
Action $ifa~x~y$ (if-food-ahead) performs action $x$ 
if a food piece is ahead of the ant,
otherwise it performs action $y$.
Action $p2~x~y$ first performs action $x$ and after it action $y$.
Action $p3~x~y~z$ first performs action $x$, 
after that action $y$ and finally $z$.
Actions $l, r$ and $m$ each take one time step to execute.
Ants action is performed over and over again until it reaches predefined
maximal number of steps. 
Fitness value is equal to number of eaten food pieces.
An individual such that eats all 89 pieces of food is 
considered as a correct solution.
This limit is set to be 600\footnote{
In \cite{koza92} this limit is said to be 400 time steps.
But there is also mentioned following solution, 
which is described as correct solution:
\texttt{(ifa m (p3 l (p2 (ifa m r) (p2 r (p2 l r)))(p2 (ifa m l) m)))}.
This program needs 545
time steps; if it is given only 400 time steps, then it eats only 79 pieces
of food. Thus we use 600 time steps. 
} time steps.

\subsection{Even parity problem}
...
\section{Conclusions}

\begin{todo}
 & do future work bych napsal, že díky jednoduchosti 
   toho co strategie dělá je to idealní kandidát na 
   optimalizaci pomocí samotnýho GP
\end{todo}
